Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE)

### 2014-02-04 WSSSPE1_report/WSSSPE1_summary_report.tex
Changes made to improve the flow in Section 3.1 Keynote of Bourne.

The first
keynote\footnote{\url{http://wssspe.researchcomputing.org.uk/keynotes/\#bourne}}
was delivered by Philip E. Bourne of UCSD. The slides are
online.\footnote{\url{www.slideshare.net/pebourne/a-recipe-for-sustainable-software}}

Philip is a basic biomedical scientist but has formed four software
companies. He also helped found PLOS and
PDB\footnote{\url{http://www.rcsb.org/pdb/home/home.do}}.

He emphasized that Sustainability for software ``does not just mean
more money from Government.''

He noted that creating effective illustrations is a key function of
software in the biomedical sciences: ``Automating Conceptualization''
(Visualizing molecular structures is a problem that has now been
solved, he is working on 3-d visualizations of the contents of cells.)

He has looked into reproducibility in computational biology, and has
concluded that ``I have proved I cannot reproduce research from my own
lab''~\cite{Veretnik}.

He noted that in places, development in science has improved thanks to
open source and Github, but for the most part remains arcane. He
argues that there is a lot we can learn from the app-store model. He
mentioned BioJava\footnote{\url{http://biojava.org/wiki/Main_Page}}
and
OpenScienceDataCloud\footnote{\url{https://www.opensciencedatacloud.org/}}
as distribution channels.

He noted a common evolutionary pathway for computational biology
projects, from data archive to analytics platform to educational
use. He suggests that use for outreach might be the final step.

He experienced problems with staff retention with respect to private
organizations which reward those combining research and software
expertise (the ``Google Bus''). However, he is a strong fan of private
sector sustainability sources and public-private partnerships. He
noted that making a successful business from scientific software alone
is very rare: founders overvalue while customers undervalue. He noted
that to last, an open source project needs a minimal funding
requirement even with a vibrant community -- good will only goes so
far if you are being paid to do something else.

He talked about grant schemes of relevance in the U.S., particularly
with regard to technology
transfer.\footnote{\url{http://www.nsf.gov/pubs/policydocs/pappguide/nsf09_1/gpg_2.jsp\#IID2}} \footnote{\url{http://www.nsf.gov/eng/iip/sbir/}}

He had problems with selling research software: the technology
transfer office in his university wanted huge IP reachthrough, whereby
they would get a share of profits from drugs developed by pharma
companies who use the software. He was aware this was unrealistic but
the technology transfer office insisted for a while. He wants to push
a one-click approach to purchasing university-written software.

@@

He then expanded on the idea of the \textbf{digital enterprise}, where
IT underpins the whole of organizational activities. He contends that
universities are way behind the curve on this. In particular, he
highlights the separation of research, teaching and administration into
silos without a common IT framework as a blocker to many useful
organizational innovations: ``University 2.0 is yet to happen.'' He
spoke of a circumstance where someone had used an algorithm developed
for computational biology in marketing.

>>>>

The first
keynote\footnote{\url{http://wssspe.researchcomputing.org.uk/keynotes/\#bourne}}
was delivered by Philip E. Bourne of UCSD. The slides are available
online.\footnote{\url{www.slideshare.net/pebourne/a-recipe-for-sustainable-software}}
Philip is a basic biomedical scientist but has formed four software
companies. He also helped found PLOS and
PDB\footnote{\url{http://www.rcsb.org/pdb/home/home.do}}.  He is
working on automating three-dimensional visualizations of cell
contents and molecular structures, a problem that has not been solved
and would serve as a key function of software in biomedical sciences.

Philip emphasized that Sustainability for software ``does not just
mean more money from Government.''  He noted that in places,
development in science has improved thanks to open source and Github,
but for the most part remains arcane. He argued that there is much we
can learn from the app-store model. He mentioned
BioJava\footnote{\url{http://biojava.org/wiki/Main_Page}} and
OpenScienceDataCloud\footnote{\url{https://www.opensciencedatacloud.org/}}
as distribution channels.  On a related note, Philip observed a common
evolutionary pathway for computational biology projects, from data
archive to analytics platform to educational use, and suggested that
use of scientific software for outreach might be the final step.

Philip shared with the audience a few real challenges he
encountered. First, he has looked into reproducibility in
computational biology, and has concluded that ``I have proved I cannot
reproduce research from my own lab''~\cite{Veretnik}.

Another problem he experienced was staff retention with respect to
private organizations which reward those combining research and
software expertise (the ``Google Bus''). However, he is a strong fan
of private sector sustainability sources and public-private
partnerships. He noted that making a successful business from
scientific software alone is very rare: founders overvalue while
customers undervalue. He noted that to last, an open source project
needs a minimal funding requirement even with a vibrant community --
good will only goes so far if you are being paid to do something else.
He talked about grant schemes of relevance in the U.S., particularly
with regard to technology
transfer.\footnote{\url{http://www.nsf.gov/pubs/policydocs/pappguide/nsf09_1/gpg_2.jsp\#IID2}} \footnote{\url{http://www.nsf.gov/eng/iip/sbir/}}

He also had problems with selling research software: the technology
transfer office in his university wanted huge IP reachthrough, whereby
they would get a share of profits from drugs developed by pharma
companies who use the software. He was aware this was unrealistic but
the technology transfer office insisted for a while. He wants to push
a one-click approach to purchasing university-written software.

@@

On the idea of the \textbf{digital enterprise}, where
IT underpins the whole of organizational activities, he contended that
universities are way behind the curve on this. In particular, he
highlighted the separation of research, teaching and administration into
silos without a common IT framework as a blocker to many useful
organizational innovations: ``University 2.0 is yet to happen.'' He
spoke of a circumstance where someone had used an algorithm developed
for computational biology in marketing.
